\vspace{30mm} {\Large\bf RISC-V Duo-PLIC for the Advancced Interrupt Architecture} \\ {\large Document Version \DuoPLICRev}

### 2023-05-09 doc/src/CSRs.tex
Upgrade chapter status; remove Duo-PLIC chapter to a separate document
Signed-off-by: John Hauser <31252952+jhauser-us@users.noreply.github.com>
Change descriptions of high-half CSRs to match latest Priv ISA
Signed-off-by: John Hauser <31252952+jhauser-us@users.noreply.github.com>

Table~\ref{tab:CSRs-M} lists the CSRs added for machine level.
Each of these registers is MXLEN bits in size.

\begin{table*}[h!]
\begin{center}
\begin{tabular}{|l|l|l|l|}
\hline
Number & Privilege & Name      & Description \\
\hline
\hline
\multicolumn{4}{|c|}{Machine-Level Window to Indirectly Accessed Registers} \\
\hline
\z{0x350} & MRW & \z{miselect} & Machine indirect register select \\
\z{0x351} & MRW & \z{mireg}    & Machine indirect register alias \\
\hline
\multicolumn{4}{|c|}{Machine-Level Interrupts} \\
\hline
\z{0x35C} & MRW & \z{mtopei}   & Machine top external interrupt
                                  (only with an IMSIC) \\
\z{0xFB0} & MRO & \z{mtopi}    & Machine top interrupt \\
\hline
\multicolumn{4}{|c|}{Virtual Interrupts for Supervisor Level} \\
\hline
\z{0x308} & MRW & \z{mvien}    & Machine virtual interrupt enables \\
\z{0x309} & MRW & \z{mvip}     & Machine virtual interrupt-pending bits \\
\hline
\multicolumn{4}{|c|}{%
  Machine-Level High-Half CSRs for Interrupts 32--63 (RV32 only)} \\
\hline
\z{0x313} & MRW & \z{midelegh} & Extension of \z{mideleg} (only with S-mode) \\
\z{0x314} & MRW & \z{mieh}     & Extension of \z{mie} \\
\z{0x318} & MRW & \z{mvienh}   & Extension of \z{mvien} (only with S-mode) \\
\z{0x319} & MRW & \z{mviph}    & Extension of \z{mvip} (only with S-mode) \\
\z{0x354} & MRW & \z{miph}     & Extension of \z{mip} \\
\hline
\end{tabular}
\end{center}
\caption{Machine-level CSRs added by the Advanced Interrupt Architecture.}
\label{tab:CSRs-M}
\end{table*}

For RV32, the \emph{high-half} CSRs listed in the table extend
registers \z{mideleg}, \z{mie}, \z{mvien}, \z{mvip}, and \z{mip} to
support a total of 64 interrupt causes.
Each low-half register (e.g.\ \z{mideleg}) covers interrupt numbers
0--31, one bit per interrupt number, while its high-half partner
(\z{midelegh}) covers interrupt numbers 32--63, again one bit per
interrupt number.
For RV64, a single \mbox{64-bit} CSR (\z{mideleg}) covers the same set
of interrupt numbers, 0--63.
The Advanced Interrupt Architecture requires that these high-half CSRs
exist for RV32, but they may all be merely read-only zeros.

@@

When \z{miselect} is a number in a reserved range (currently
\z{0x00}--\z{0x2F}, \z{0x40}--\z{0x6F}, or a number above \z{0xFF}
not designated for custom use), attempts to access \z{mireg} raise an
illegal instruction exception.

@@

When \mbox{S-mode} is implemented, CSRs \z{mvien} and \z{mvip}
(together with their high-half partners for RV32) support interrupt
filtering and virtual interrupts for supervisor level.
These facilities are explained in Section~\ref{sec:virtIntrs-S}.

@@

Table~\ref{tab:CSRs-S} lists the supervisor-level CSRs that are added
if the hart implements \mbox{S-mode}.
Each of these registers is SXLEN bits in size.
The functions of these registers all match their machine-level
counterparts.

\begin{table*}[h!]
\begin{center}
\begin{tabular}{|l|l|l|l|}
\hline
Number & Privilege & Name      & Description \\
\hline
\hline
\multicolumn{4}{|c|}{%
  Supervisor-Level Window to Indirectly Accessed Registers} \\
\hline
\z{0x150} & SRW & \z{siselect} & Supervisor indirect register select \\
\z{0x151} & SRW & \z{sireg}    & Supervisor indirect register alias \\
\hline
\multicolumn{4}{|c|}{Supervisor-Level Interrupts} \\
\hline
\z{0x15C} & SRW & \z{stopei}   & Supervisor top external interrupt
                                  (only with an IMSIC) \\
\z{0xDB0} & SRO & \z{stopi}    & Supervisor top interrupt \\
\hline
\multicolumn{4}{|c|}{%
  Supervisor-Level High-Half CSRs for Interrupts 32--63 (RV32 only)} \\
\hline
\z{0x114} & SRW & \z{sieh}     & Extension of \z{sie} \\
\z{0x154} & SRW & \z{siph}     & Extension of \z{sip} \\
\hline
\end{tabular}
\end{center}
\caption{Supervisor-level CSRs added by the Advanced Interrupt Architecture.}
\label{tab:CSRs-S}
\end{table*}

@@

When \z{siselect} is a number in a reserved range (currently
\z{0x00}--\z{0x2F}, \z{0x40}--\z{0x6F}, or a number above \z{0xFF}
not designated for custom use), or in the range \z{0x70}--\z{0xFF}
when there is no IMSIC, attempts to access \z{sireg} raise an illegal
instruction exception (unless executing in a virtual machine, covered
in the next section).

@@

If a hart implements the Privileged Architecture's hypervisor
extension, then the hypervisor and VS CSRs listed in
Table~\ref{tab:CSRs-hypervisor} are also added.
The VS CSRs are all VSXLEN bits wide,
while the hypervisor CSRs are all HSXLEN bits.
(\textit{HSXLEN} is just another name for
SXLEN for hypervisor-extended \mbox{S-mode}).

\begin{table*}[h!]
\begin{center}
\begin{tabular}{|l|l|l|l|}
\hline
Number & Privilege & Name        & Description \\
\hline
\hline
\multicolumn{4}{|c|}{%
  Virtual Interrupts and Interrupt Priorities for VS Level} \\
\hline
\z{0x608} & HRW & \z{hvien}      & Hypervisor virtual interrupt enables \\
\z{0x609} & HRW & \z{hvictl}     & Hypervisor virtual interrupt control \\
\z{0x646} & HRW & \z{hviprio1}   & Hypervisor VS-level interrupt priorities \\
\z{0x647} & HRW & \z{hviprio2}   & Hypervisor VS-level interrupt priorities \\
\hline
\multicolumn{4}{|c|}{VS-Level Window to Indirectly Accessed Registers} \\
\hline
\z{0x250} & HRW & \z{vsiselect}
                   & Virtual supervisor indirect register select \\
\z{0x251} & HRW & \z{vsireg}
                   & Virtual supervisor indirect register alias \\
\hline
\multicolumn{4}{|c|}{VS-Level Interrupts} \\
\hline
\z{0x25C} & HRW & \z{vstopei}    & Virtual supervisor top external interrupt
                                    (only with an IMSIC) \\
\z{0xEB0} & HRO & \z{vstopi}     & Virtual supervisor top interrupt \\
\hline
\multicolumn{4}{|c|}{Hypervisor and VS-Level High-Half CSRs (RV32 only)} \\
\hline
\z{0x613} & HRW & \z{hidelegh}   & Extension of \z{hideleg} \\
\z{0x618} & HRW & \z{hvienh}     & Extension of \z{hvien} \\
\z{0x655} & HRW & \z{hviph}      & Extension of \z{hvip} \\
\z{0x656} & HRW & \z{hviprio1h}  & Extension of \z{hviprio1} \\
\z{0x657} & HRW & \z{hviprio2h}  & Extension of \z{hviprio2} \\
\z{0x214} & HRW & \z{vsieh}      & Extension of \z{vsie} \\
\z{0x254} & HRW & \z{vsiph}      & Extension of \z{vsip} \\
\hline
\end{tabular}
\end{center}
\caption{Hypervisor and VS CSRs added by the Advanced Interrupt Architecture.}
\label{tab:CSRs-hypervisor}
\end{table*}

The VS CSRs in the table (\z{vsiselect}, \z{vsireg},
\z{vstopei}, \z{vstopi}, and, for RV32, \z{vsieh} and \z{vsiph})
all match supervisor CSRs, and substitute for those
supervisor CSRs when executing in a virtual machine (in \mbox{VS-mode}
or \mbox{VU-mode}).

@@

The space of registers selectable by \z{vsiselect} is more limited than
for machine and supervisor levels:
\begin{displayLinesTable}[l@{\quad}l]
\z{0x000}--\z{0x06F} & inaccessible \\
\z{0x070}--\z{0x0FF} & external interrupts (IMSIC only), or inaccessible \\
\z{0x100}--\z{0x1FF} & inaccessible \\
\end{displayLinesTable}
When \z{vsiselect} has the number of an \emph{inaccessible} register,
including values above \z{0x1FF} without the most-significant bit set,
attempts from \mbox{M-mode} or \mbox{HS-mode} to access \z{vsireg}
raise an illegal instruction exception, and attempts from
\mbox{VS-mode} to access \z{sireg} (really \z{vsireg}) raise a virtual
instruction exception.

\begin{commentary}
Requiring a range of\/ {\rm 0}--\z{0x1FF} for \z{vsiselect}, even
though most or all of the space is inaccessible, permits a hypervisor
to emulate indirectly accessed registers in the implemented range,
including registers that may be standardized in the future at locations
\z{0x100}--\z{0x1FF}.
\end{commentary}

The locations of registers for external interrupts (numbers
\z{0x70}--\z{0xFF}) are accessible only when field VGEIN of \z{hstatus}
is the number of an implemented guest external interrupt, not zero.
If VGEIN is not the number of an implemented guest external interrupt
(including the case when no IMSIC is implemented), then all non-custom
values of \z{vsiselect} designate an inaccessible register.

@@

Aside from the high-half CSR \z{hidelegh}, the hypervisor CSRs added
by the Advanced Interrupt Architecture (\z{hvien}, \z{hvictl},
\z{hviprio1}, and \z{hviprio2}, plus the high-half hypervisor CSRs for
RV32) augment \z{hvip} for injecting interrupts into VS level.
The use of these registers is covered in Chapter~\ref{ch:VSLevel} on
interrupts for virtual machines.

@@

If the hypervisor extension is implemented, the same three bits are
defined also in \z{hstateen0} but concern only the state potentially
accessible to a virtual machine executing in privilege modes VS and VU:
\begin{displayLinesTable}[l@{\quad}l]
bit 60 & CSRs \z{siselect} and \z{sireg}
          (really \z{vsiselect} and \z{vsireg}) \\
bit 59 & CSRs \z{siph} and \z{sieh} (RV32 only) and \z{stopi}
          (really \z{vsiph}, \z{vsieh}, and \z{vstopi}) \\
bit 58 & all state of IMSIC guest interrupt files,
          including CSR \z{stopei} (really \z{vstopei}) \\
\end{displayLinesTable}

If one of these bits is zero in \z{hstateen0}, and the same bit is one
in \z{mstateen0}, then an attempt to access the corresponding state
from \mbox{VS-mode} raises a virtual instruction exception.

Bit~58 is implemented in \z{hstateen0} only if the hart has an IMSIC.
Furthermore, even with an IMSIC, bit~58 may (or may not) be read-only
zero in \z{hstateen0} if the IMSIC has no \emph{guest interrupt files}
for guest external interrupts (Chapter~\ref{ch:IMSIC}).
When this bit is zero, a virtual machine is prevented from
accessing the hart's IMSIC the same as when \z{hstatus}.VGEIN =~0.

Extension Ssstateen is defined as
the supervisor-level view of Smstateen.
Ssstateen therefore incorporates the bits defined above
for \z{hstateen0} but not those for \z{mstateen0},
since machine-level CSRs are not visible to supervisor level.

>>>>

Table~\ref{tab:CSRs-M} lists both the CSRs added for machine level
and existing machine-level CSRs whose size is changed
by the Advanced Interrupt Architecture.
Existing CSRs \z{mie}, \z{mip}, and \z{mideleg} are
widended to 64~bits to support a total of 64 interrupt causes.

\begin{table*}[h!]
\begin{center}
\begin{tabular}{|c|c|c|l|l|}
\hline
Number & Privilege & Width & Name & Description \\
\hline
\hline
\multicolumn{5}{|c|}{Machine-Level Window to Indirectly Accessed Registers} \\
\hline
\z{0x350} & MRW & MXLEN & \z{miselect} & Machine indirect register select \\
\z{0x351} & MRW & MXLEN & \z{mireg}    & Machine indirect register alias \\
\hline
\multicolumn{5}{|c|}{Machine-Level Interrupts} \\
\hline
\z{0x304} & MRW & 64    & \z{mie}    & Machine interrupt-enable bits \\
\z{0x344} & MRW & 64    & \z{mip}    & Machine interrupt-pending bits \\
\z{0x35C} & MRW & MXLEN & \z{mtopei}
                              & Machine top external interrupt (only with an \\
          &     &       &            & \quad IMSIC) \\
\z{0xFB0} & MRO & MXLEN & \z{mtopi}  & Machine top interrupt \\
\hline
\multicolumn{5}{|c|}{Delegated and Virtual Interrupts for Supervisor Level} \\
\hline
\z{0x303} & MRW & 64 & \z{mideleg} & Machine interrupt delegation \\
\z{0x308} & MRW & 64 & \z{mvien}   & Machine virtual interrupt enables \\
\z{0x309} & MRW & 64 & \z{mvip}    & Machine virtual interrupt-pending bits \\
\hline
\multicolumn{5}{|c|}{Machine-Level High-Half CSRs (RV32 only)} \\
\hline
\z{0x313} & MRW & 32 & \z{midelegh}
                        & Upper 32 bits of of \z{mideleg} (only with S-mode) \\
\z{0x314} & MRW & 32 & \z{mieh} & Upper 32 bits of \z{mie} \\
\z{0x318} & MRW & 32 & \z{mvienh}
                             & Upper 32 bits of \z{mvien} (only with S-mode) \\
\z{0x319} & MRW & 32 & \z{mviph}
                              & Upper 32 bits of \z{mvip} (only with S-mode) \\
\z{0x354} & MRW & 32 & \z{miph} & Upper 32 bits of \z{mip} \\
\hline
\end{tabular}
\end{center}
\caption{%
Machine-level CSRs added or widened by the Advanced Interrupt Architecture.%
}
\label{tab:CSRs-M}
\end{table*}

For RV32, the \emph{high-half} CSRs listed in the table
allow access to the upper 32~bits of
registers \z{mideleg}, \z{mie}, \z{mvien}, \z{mvip}, and \z{mip}.
The Advanced Interrupt Architecture requires that these high-half CSRs
exist for RV32, but the bits they access may all be merely read-only zeros.

@@

When \z{miselect} is a number in a reserved range (currently
\z{0x00}--\z{0x2F}, \z{0x40}--\z{0x6F}, or a number above \z{0xFF}
not designated for custom use), attempts to access \z{mireg}
will typically raise an
illegal instruction exception.

@@

When \mbox{S-mode} is implemented, CSRs \z{mvien} and \z{mvip}
support interrupt
filtering and virtual interrupts for supervisor level.
These facilities are explained in Section~\ref{sec:virtIntrs-S}.

@@

Table~\ref{tab:CSRs-S} lists the supervisor-level CSRs that are added
and existing CSRs that are widened to 64~bits,
if the hart implements \mbox{S-mode}.
The functions of these registers all match their machine-level
counterparts.

\begin{table*}[h!]
\begin{center}
\begin{tabular}{|c|c|c|l|l|}
\hline
Number & Privilege & Width & Name      & Description \\
\hline
\hline
\multicolumn{5}{|c|}{%
  Supervisor-Level Window to Indirectly Accessed Registers} \\
\hline
\z{0x150} & SRW & SXLEN & \z{siselect} & Supervisor indirect register select \\
\z{0x151} & SRW & SXLEN & \z{sireg}    & Supervisor indirect register alias \\
\hline
\multicolumn{5}{|c|}{Supervisor-Level Interrupts} \\
\hline
\z{0x104} & SRW & 64    & \z{sie}      & Supervisor interrupt-enable bits \\
\z{0x144} & SRW & 64    & \z{sip}      & Supervisor interrupt-pending bits \\
\z{0x15C} & SRW & SXLEN & \z{stopei}
                                   & Supervisor top external interrupt (only \\
          &     &       &          & \quad with an IMSIC) \\
\z{0xDB0} & SRO & SXLEN & \z{stopi}    & Supervisor top interrupt \\
\hline
\multicolumn{5}{|c|}{Supervisor-Level High-Half CSRs (RV32 only)} \\
\hline
\z{0x114} & SRW & 32    & \z{sieh}     & Upper 32 bits of \z{sie} \\
\z{0x154} & SRW & 32    & \z{siph}     & Upper 32 bits of \z{sip} \\
\hline
\end{tabular}
\end{center}
\caption{%
Supervisor-level CSRs added or widened by the Advanced Interrupt Architecture.%
}
\label{tab:CSRs-S}
\end{table*}

@@

When \z{siselect} is a number in a reserved range (currently
\z{0x00}--\z{0x2F}, \z{0x40}--\z{0x6F}, or a number above \z{0xFF}
not designated for custom use), or in the range \z{0x70}--\z{0xFF}
when there is no IMSIC, attempts to access \z{sireg}
should preferably raise an illegal
instruction exception (unless executing in a virtual machine, covered
in the next section).

@@

If a hart implements the Privileged Architecture's hypervisor
extension, then the hypervisor and VS CSRs listed in
Table~\ref{tab:CSRs-hypervisor} are also either added
or widened to 64~bits.

\begin{table*}[h!]
\begin{center}
\begin{tabular}{|c|c|c|l|l|}
\hline
Number & Privilege & Width & Name & Description \\
\hline
\hline
\multicolumn{5}{|c|}{%
  Delegated and Virtual Interrupts, Interrupt Priorities, for VS Level%
} \\
\hline
\z{0x603} & HRW & 64     & \z{hideleg} & Hypervisor interrupt delegation \\
\z{0x608} & HRW & 64     & \z{hvien}  & Hypervisor virtual interrupt enables \\
\z{0x609} & HRW & HSXLEN & \z{hvictl} & Hypervisor virtual interrupt control \\
\z{0x645} & HRW & 64     & \z{hvip}
                                 & Hypervisor virtual interrupt-pending bits \\
\z{0x646} & HRW & 64     & \z{hviprio1}
                                 & Hypervisor VS-level interrupt priorities \\
\z{0x647} & HRW & 64     & \z{hviprio2}
                                 & Hypervisor VS-level interrupt priorities \\
\hline
\multicolumn{5}{|c|}{VS-Level Window to Indirectly Accessed Registers} \\
\hline
\z{0x250} & HRW & VSXLEN & \z{vsiselect}
                               & Virtual supervisor indirect register select \\
\z{0x251} & HRW & VSXLEN & \z{vsireg}
                               & Virtual supervisor indirect register alias \\
\hline
\multicolumn{5}{|c|}{VS-Level Interrupts} \\
\hline
\z{0x204} & HRW & 64     & \z{vsie}
                                 & Virtual supervisor interrupt-enable bits \\
\z{0x244} & HRW & 64     & \z{vsip}
                                 & Virtual supervisor interrupt-pending bits \\
\z{0x25C} & HRW & VSXLEN & \z{vstopei}
                           & Virtual supervisor top external interrupt (only \\
          &     &        & & \quad with an IMSIC) \\
\z{0xEB0} & HRO & VSXLEN & \z{vstopi} & Virtual supervisor top interrupt \\
\hline
\multicolumn{5}{|c|}{Hypervisor and VS-Level High-Half CSRs (RV32 only)} \\
\hline
\z{0x613} & HRW & 32     & \z{hidelegh}  & Upper 32 bits of \z{hideleg} \\
\z{0x618} & HRW & 32     & \z{hvienh}    & Upper 32 bits of \z{hvien} \\
\z{0x655} & HRW & 32     & \z{hviph}     & Upper 32 bits of \z{hvip} \\
\z{0x656} & HRW & 32     & \z{hviprio1h} & Upper 32 bits of \z{hviprio1} \\
\z{0x657} & HRW & 32     & \z{hviprio2h} & Upper 32 bits of \z{hviprio2} \\
\z{0x214} & HRW & 32     & \z{vsieh}     & Upper 32 bits of \z{vsie} \\
\z{0x254} & HRW & 32     & \z{vsiph}     & Upper 32 bits of \z{vsip} \\
\hline
\end{tabular}
\end{center}
\caption{%
Hypervisor and VS CSRs added or widened
by the Advanced Interrupt Architecture.
(Parameter HSXLEN is just another name for
SXLEN for hypervisor-extended \mbox{S-mode}).%
}
\label{tab:CSRs-hypervisor}
\end{table*}

The new VS CSRs in the table (\z{vsiselect}, \z{vsireg},
\z{vstopei}, and \z{vstopi})
all match supervisor CSRs, and substitute for those
supervisor CSRs when executing in a virtual machine (in \mbox{VS-mode}
or \mbox{VU-mode}).

@@

The space of registers selectable by \z{vsiselect} is more limited than
for machine and supervisor levels:
\begin{displayLinesTable}[l@{\quad}l]
\z{0x000}--\z{0x02F} & reserved \\
\z{0x030}--\z{0x03F} & inaccessible \\
\z{0x040}--\z{0x06F} & reserved \\
\z{0x070}--\z{0x0FF} & external interrupts (IMSIC only), or inaccessible \\
\z{0x100}--\z{0x1FF} & reserved \\
\end{displayLinesTable}
When \z{vsiselect} has the number of an \emph{inaccessible} register,
attempts from \mbox{M-mode} or \mbox{HS-mode} to access \z{vsireg}
raise an illegal instruction exception, and attempts from
\mbox{VS-mode} or \mbox{VU-mode} to access \z{sireg}
(really \z{vsireg}) raise a virtual instruction exception.
Similarly, when \z{vsiselect} has a reserved value,
including values above \z{0x1FF} without the most-significant bit set,
attempts from \mbox{M-mode} or \mbox{HS-mode} to access \z{vsireg}
should preferably raise an illegal instruction exception, and attempts from
\mbox{VS-mode} or \mbox{VU-mode} to access \z{sireg} (really \z{vsireg})
should preferably raise a virtual instruction exception.

\begin{commentary}
Requiring a range of\/ {\rm 0}--\z{0x1FF} for \z{vsiselect}, even
though most or all of the space is
reserved or inaccessible, permits a hypervisor
to emulate indirectly accessed registers in the implemented range,
including registers that may be standardized in the future at locations
\z{0x100}--\z{0x1FF}.
\end{commentary}

The locations of registers for external interrupts (numbers
\z{0x70}--\z{0xFF}) are accessible only when field VGEIN of \z{hstatus}
is the number of an implemented guest external interrupt, not zero.
If VGEIN is not the number of an implemented guest external interrupt
(including the case when no IMSIC is implemented), then all non-custom
values of \z{vsiselect} are either reserved
or designate an inaccessible register.

@@

The new hypervisor CSRs in Table~\ref{tab:CSRs-hypervisor}
(\z{hvien}, \z{hvictl}, \z{hviprio1}, and \z{hviprio2})
augment \z{hvip} for injecting interrupts into VS level.
The use of these registers is covered in Chapter~\ref{ch:VSLevel} on
interrupts for virtual machines.

@@

If the hypervisor extension is implemented, the same three bits are
defined also in hypservisor CSR \z{hstateen0}
but concern only the state potentially
accessible to a virtual machine executing in privilege modes VS and VU:
\begin{displayLinesTable}[l@{\quad}l]
bit 60 & CSRs \z{siselect} and \z{sireg}
          (really \z{vsiselect} and \z{vsireg}) \\
bit 59 & CSRs \z{siph} and \z{sieh} (RV32 only) and \z{stopi}
          (really \z{vsiph}, \z{vsieh}, and \z{vstopi}) \\
bit 58 & all state of IMSIC guest interrupt files,
          including CSR \z{stopei} (really \z{vstopei}) \\
\end{displayLinesTable}

If one of these bits is zero in \z{hstateen0},
and the same bit is one in \z{mstateen0},
then an attempt to access the corresponding state
from VS or \mbox{VU-mode} raises a virtual instruction exception.
(But note that, for high-half CSRs \z{siph} and
\z{sieh}, this applies only when XLEN =~32.
When $\mbox{XLEN} > \mbox{32}$, an attempt to access
\z{siph} or \z{sieh} raises an illegal instruction
exception as usual, not a virtual instruction exception.)

Bit~58 is implemented in \z{hstateen0}
only if the hart has an IMSIC.
Furthermore, even with an IMSIC, bit~58 may (or may not) be read-only
zero in \z{hstateen0} if
the IMSIC has no \emph{guest interrupt files}
for guest external interrupts (Chapter~\ref{ch:IMSIC}).
When this bit is zero (whether read-only zero or set to zero),
a virtual machine is prevented from
accessing the hart's IMSIC the same as when \z{hstatus}.VGEIN =~0.

Extension Ssstateen is defined as
the supervisor-level view of Smstateen.
Therefore, the combination of Ssaia and Ssstateen incorporates the bits
defined above for \z{hstateen0} but not those for \z{mstateen0},
since machine-level CSRs are not visible to supervisor level.
